---
title: 'Topic 1: Introduction to R'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# September 2018 report

###

#### Vectors

Vector is a sequence of data elements of the same type.  We can have character vector, numeric vectors, logical vector. To create a vector, we use **c()**. 

Now, we create a character vector

```{r}
programming_languages = c("C++", "R",  "Scala","Python", "JavaScript")
# print it
programming_languages
```

Here, **programming_languages** is a variable name and to the right of **=** are the values of this variable. Note, R considers **programming_languages** as a character vector.

If we create a a single value and assign to any variable, that variable is considered as a vector.

```{r}
income = 30000
is.vector(income)
```

We see TRUE, that is, price is a vector.


#### Integer vector

Now, we will make an integer vector of prices of common foods. We will make an integer vector **price** consisting of prices of sugar, rice, potato and dry fish.

```{r}
price = c(60, 70, 30, 600)
# print it
price
```

We could also label the values with the food names. We need to use **names(vector)** for this purpose. ow, we create a vector of food names.

```{r}
nam = c("sugar", "rice", "potato", "dry fish")
nam
```

Ler's label our variable **food.price** using **names()**.

```{r}
names(price) = nam
# Print
price
#nam
```

We can do all of these in one line.

```{r}
price2 = c(sugar = 60, rice = 70, potato  = 30, "dry fish" = 600)
price2
``` 

We can take a look at the structure of an R object using **str(object)**

```{r}
str(price2)
```

#### Vector arithmetic
 
We can do mathematical operations such as addition, subtraction, multiplication, division etc. with vectors

```{r}
item1 = 700
item2 = 650
total = item1 + item2
total
```

#### Element-wise opearation or vectorized operation

```{r}
a = c(10, 20, 30)
b = c(1,2,3)
a + b
```


#### Subset by index

We can access the second element of price2 vector using [] or indexing in the following way:

```{r}
price2 = c(sugar = 60, rice = 70, potato  = 30, "dry fish" = 600)
#price2
price2[1]
```


We can also get the value of the first item by indexing by its name

```{r}
price2["sugar"]
```

#### Exercise: Get the second item of price2 in 2 different ways. Give a try.

We can select multiple elements using c() in this way: **c(element_number1, element_number2)** inside **[]**

```{r}
price2[c(1,2)]
```

We can also use names to select multiple elements.

```{r}
price2[c("sugar", "rice")]
```

#### Selecting with deselecting or without selecting particular value or values

Use **-** as deselectg operator

We can use **-** before **c(element_number, ...)**

Let's select all but the third price

```{r}
price2[-1]
```

#### Deselecting multiple items

Now, we select all but first and third element, we just add a **-** before **c(first element, third element)** 

```{r}
price2
price2[-c(1, 3)]
```

But, we can't use name similarly as positional index to deselect items. So, **price2[-c("sugar", "potato")]**, will not work.

### Using Logical Vector for subsetting

```{r}
price2[c(FALSE, FALSE, TRUE, TRUE)]
price2
```

### Matrix -- matrix()

Matrix is a 2 dimensional array of elements with fixed number of rows and columns. It can contain only one type of element; either numeric or character, or other type and it doesn't allow a mix of different types of data elements.

We create a matrix using **matrix()** command. We can specify the number of rows using **nrow** and number of columns using **ncol** argument.

```{r}
number = matrix(1:25, nrow = 5)
# Print this matrix
number
```

Let's investigate what we have just done. **matrix()** tells that it is a matrix, **1:25** tellss that the matrix consists of numbers from 1 to 25 and **nrow = 5** tells that there will be 5 rows and these numbers have to be arranged column-wise.  argument.

```{r}
mat2 = matrix(1:12, nrow = 4)
mat2
```

The values here will be arranged columnwise.

If we want the values to be filled rowwise. We will use **byrow** argument for doing it.

```{r}
mat3 = matrix(1:12, nrow = 4, byrow = TRUE)
mat3
```

### Naming a matrix using rownames() and colnames()

```{r}
matrix_arbitrary = matrix(1:16, nrow = 4, byrow = TRUE, dimnames = list(c("row1", "row2", "row3", "row4"), 
                                                             c("col1", "col2", "col3", "col4")))
matrix_arbitrary
```

### Categorical variables as factors -- factor()

We can use **factor()** to convert characters to factor. Please take note that R orders factor levels alphabetically.

```{r}
quality = c("bad", "good", "very bad", "very good", "medium")
quality = factor(quality)
quality
```

Let's specify the order of levels using **levels** argument inside **factor()**.

```{r}
quality2 = factor(quality, levels = c("very bad", "bad", "medium", "good", "very good"))
quality2
str(quality2)
```

We can also rename the levels in the following ways

```{r}
levels(quality2) = c("VB", "B", "M", "G", "VG")
quality2
# or we can do the following
quality3 = factor(quality, labels = c("VB", "B", "M", "G", "VG"))
quality
```

For making Ordered factors, use **ordered = TRUE** argument

```{r}
quality = c("bad", "good", "very bad", "very good", "medium")
quality2 = factor(quality, ordered = TRUE,
                   levels = c("very bad", "bad", "medium", "good", "very good"))
quality2
quality2[1] < quality2[2]
```


### Working with list --  list()

#### List allows a variable to contain all types of R objects or a mix of different types of objects.

Example --  Suppose, we want to group different employee information. We now use vector in the folllowing way:

```{r}
staff = c("Sohel Parvez", 2015, "Accountant", "Excellent")
```

In this case, as vector doesn't permit mixing of different data types, the integer element is coerced to character which we don't want. Using list, we can have preserve data types of different elements.

```{r}
staff = list("Sohel Parvez", 2016, "Accountant", "Excellent")
staff
```

Now we put label on the staff variable.

```{r}
names(staff) = c("Name", "Start_year", "Position", "Performance")
```

Like vectors, we can also do these in one line in the following way.

```{r}
staff = list(name = "Sohel Parvez", start_year = 2016, position = "Accountant", performance = "Excellent")
```

List can be nested. Use **str()** to print the elements of a list in a compact way.

#### Accessing list element or indexing list.

```{r}
staff[1]
staff
```

We get the element as a list and not as a character string. To get the character string of names, we need to use **[[]]**. Here is how it looks like. 

```{r}
staff[[1]]
```

We can also get multiple elements as a list.

```{r}
staff[c(1,2)]
```

Now, can we select mulltiple elements as character strings and not as list? NO! It throws error.

**staff[[c(1,2)]**

It might work when we have list within a list. Let's define another employee.

```{r}
staff_new = list(name = "Andrew Ng", Start_year = 2009, position = "Chief of AI", performance = "Excellent", staff)
staff_new
second_staff_name = staff_new[[c(5,1)]]
# We can also do
second_staff_name = staff_new[[5]][[1]]
```

We can also subset by name.

```{r}
# Find the name of the first employee
staff_new[["name"]]
```

```{r}
# We can also use single bracket, returns a list
staff_new["name"] 
```

We can also select subset using logicals. Logical works with single bracket **[]**

```{r}
staff_new[c(TRUE,  TRUE, TRUE, FALSE, FALSE)]
```

But logical doesn't work with multiple brackets. **staff_new[[c(TRUE, FALSE, TRUE, FALSE, FALSE)]]** throws error. 
We can also use **$** to select element

```{r}
staff_new$name
```

Using **$**, we can also add element to the list

```{r}
salary = c(80000, 6000000)
staff_new$salary = salary
staff_new
```

We also could do the same thing in the following way:

```{r}
staff_new[["salary"]] = salary
```

Suppose we have the following three lists providing information of educational qualification, address and salary of employees.

```{r}
edu = c("Masters", "PhD", "Masters", "BSc", "BBA")
address = c("Shewrapara", "Baridhara", "Gulshan 2", "Mirpur 1", "Uttara")
salary = c(50000, 100000, 50000, 30000, 20000, 25000)
employee_info = list(edu, address, salary)
```

Now, if we want the information of salary of second employee and address of fifth employee, we can do the following.

```{r}
employee_info[[3]][2]
employee_info[[2]][5]
```

### Data Frames

We create data frame  using **data.frame()**

Suppose, we have conducted a survey and have information on *name*, *age*, *income* and *spending*. We will create a data frame information.

```{r}
name = c("Mustafa", "Mintu", "Ikram", "Shahadat", "Hakim", "Rony")
age = c(30, 24, 34, 45, 32, 52)
income = c(60000, 56000, 310000, 435600, 30000, 90500)
spending = c(50000, 45000, 200000, 300000, 28000, 77000)
information = data.frame(name, age, income, spending)
information
```

We can also name data frame using **names()**.

```{r}
names(information) = c("Name", "Age", "Income", "Spending")
information
```

We can also name data frames in the following way:

```{r}
information = data.frame(Name = name, Age = age, Income = income, Spending = spending)
information
str(information)
```

We can see that *R* is considering *Name* as factor variable. We can change this behavior by setting **stringsAsFact0rs = FALSE**

```{r}
information = data.frame(Name = name, Age = age, Income = income, Spending = spending, 
                         stringsAsFactors = FALSE)
information
# Now Name will be character
str(information)
```

To have a look at the first few observations of a data frame, use **head()** function.

```{r}
head(information)
```

To know the number of rows and columns of a data frame, we use **dim()** function

```{r}
dim(information)
```

We can subset data frame using [] (as in matrices) and also [[]] and $ (as in list). To select second row and third column, we write

```{r}
information[2, 3]
```

We can also do this

```{r}
information[2, "Income"]
information[, "Age"]
```

We can also select 2nd and 4th row and **Income** and **Age** at the same time

```{r}
information[c(2, 4), c("Income", "Age")]
```

We can also use **$**  and [[]] like below to select. In all of the following cases we get vector. 

```{r}
information$Income
# or
information[["Income"]]
# or we can also use index
information[[3]]
```

But if we use **[]**, we will get a data frame or a list.

```{r}
information["Income"]
```

Now we will look at how we can add columns and rows.

### Adding column

The use of **$** and [[]] has the same effect of adding columns. Let's suppose that we want to add column books to inform us how many books of the office employees owe.

```{r}
books = c(10, 12, 5, 2, 4, 3)
# Now add books to information dataframe
information$books = books
# or we can use
information[["books"]] = books
information
```

We can also use cbind() to add column to a dataframe.

```{r}
pencils = c(2, 4, 6, 1, 0, 3)
information = cbind(information, pencils)
information
```

Similarly, we can also use rbind(). We just figured out that we have left out data for another employee Bulbul. We want to store his information in the information dataframe.

```{r}
# Note how we have to name the elements and have to use data.frame and not c()
bulbul = data.frame(Name = "Bulbul", Age = 40, Income = 115000, Spending = 90000, books = 5, 
                    pencils = 4)
information = rbind(information, bulbul)
information
```

## Basic Plots

Let's generate some random variables first and make a scatter plot of these.

```{r, fig.height=2.5}
x = rnorm(100)
y = rnorm(100)
plot(x, y)
```

Now we can modify this plot to have different colour by using **pch** for differnt plot sybbols and **col** for different plot colours

```{r, fig.height=2.5}
plot(x, y, pch = 19, col = 'red')
```

We can make a line plot using **type = 'l'** in **plot()** function
```{r, fig.height=2.5}
x2 = seq(0, 2*pi, len = 100)
y2 = sin(x2)
plot(x2, y2, type = "l")
```

We can further modify this plot by increasing line width using **lwd** and **col**

```{r, fig.height=2.5}
plot(x2, y2, type = "l", col = "darkgreen")
```

Now we can further modify this plot by including points using **points** and giving the limit of y-axis by using **ylim=c(lower bound, upper bound)**

**col.main** specifies the color of the title. **cex.axis** determines the font size of axis labels.

```{r, fig.height=2.5}
plot(x2, y2, type = "l", col = "darkblue", lwd = 3, ylim = c(-1.2, 1.2), xlab = "x-axis title", 
     ylab = "y-axis title", col.main = "Orange", main = "Title of the graph", cex.axis = 0.8, 
     sub = "Sub title")
# Now we will add values from a normal distribution with mean 0 and variance 0.5
y3= y2 + rnorm(100, 0, 0.5)
# Now we can add these points to the line plot of y2 versus x2
points(x2, y3, pch = 16, col = "darkred")
```

We can also get line chart and scatter plot in the same garphic by making a scatter plot using **plot()** first and then by writing **lines(x, y)** function. Just remember that **lty = 1** gives a solid line and **lty = 2** gives a dashed line. Below is an example:

```{r, fig.height=2.5}
y4 = sin(x2)
plot(x2, y2, type = "l",  col = "darkgreen")
lines(x2, y4, lwd = 3, lty = 2, col = "darkblue")
```

> Note that here **lines()** add line to an existing plot

We can also group multiple plots in a single figure using **par()**. Inside **par()**, use **mfrow = c(1,2)** to get a figure of 1 row and 2 columns. Here the fist value in *c(1,2)* in **mfrow = c(1,2)** corresponds to number of rows and the second corresponds to the number of columns. So, in other words, it is **mfrow = c(numberOfRows, numberOfColumns)**

Let's have a look at the below example

```{r, fig.height=5}
# The below command will give us 1 row and 3 columns of plots in a single figure
par(mfrow = c(1,3))
plot(x, y, pch = 19, col = 'red')
plot(x2, y2, type = "l",  col = "darkgreen")
plot(x2, y4, type = "l",  col = "darkred")
# Now, set to the default of 1 row and 1 column of plot in a figure
par(mfrow = c(1,1))
```

### # To reset the grid:

```{r}
#layout(1)
# Or
par(mfrow = c(1, 1))
```
