---
title: "dplyr"
author: "Shammunul Islam"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data management in dplyr

dplyr is a package develped by Hadley Wickham for easy data management, data aggregation and data summary.  

But before we get ourself introduced with dplyr, we need to become familiarized with pipe operator **%>%**

#### %>%

To the right of this operator, function is written and to the left of the operator, the first argument to the function is written.

Let's look at example code to clearly understand.

```{r}
library(dplyr)
library(hflights)
hflights %>% head()


```

```{r}
hflights %>%
  glimpse()
```

We will now be using pipe operator all throughout this module. 

We will be focusing now on 5 functions of dplyr, starting with **select**

### select

This is used to select multiple columns or single column of a dataframe.

```{r}
hflights %>% 
  select(DepTime, ArrTime) %>%
  glimpse()
```

To select the first 10 rows of this, we can also use **head()** like below:

```{r}
selected = hflights %>% 
  select(DepTime, ArrTime) %>%
  head(10)
write.csv(selected,"selected.csv")
```

We can also save this as following:

```{r}
new = hflights %>% 
  select(DepTime, ArrTime) %>%
  head(10)
new
```

Note that, to see the contents o this dataframe, we had to write the name of this after we have created **new**. We can create and see the contents of a variable by wrapping it around by **()** in the following way:

```{r}
(new = hflights %>% 
  select(DepTime, ArrTime) %>%
  head(10))
```

We can also choose columns by their positional index. Suppose, we want to select columns 2 to 5 and columns 8 to 12.

```{r}
hflights %>%
  select(2:5, 8:12)
```

### mutate

We use this function to add new column based on available columns.

Let's build a new dataframe sales now.

```{r}
sales = data.frame(area = rep(c("Syl", "Dhk", "Com", "Fen", "Khu", "Cht"), 10), dept = rep(c("A", "B", "C", "D", "E", "F"), each = 10), sale = rnorm(60, 100000, 20000), cost = rnorm(60, 30000, 10000))
glimpse(sales)
```

Now, let's create profit column using sales data.

```{r}
sales = sales %>%
  mutate(profit = sale - cost)
sales
```

### Filter -- filter

We can select data conditionally where data that meet the condition are only used or returned.

```{r}
sales %>%
  filter(dept == "B")
```

```{r}
sales %>%
  filter(cost > 40000)
```

### arrange

It rearranges dataset based on the column specified.

```{r}
sales %>%
  arrange(area + dept)
```

### group_by

Using this, we can group our data

```{r}
sales %>%
  group_by(area, dept)
```

### summarize -- gives summary statistic

```{r}
sales %>%
  summarize(minimum = min(sale), maximum = max(sale))
```

Now, suppose we want to know the min and max for different areas

```{r}
sales %>%
  group_by(area) %>%
  filter(area == 'Dhk') %>%
  summarize(minimum = min(sale), maximum = max(sale))
```

Now, you want to get a summary by both area and by department

```{r}
sales %>%
  group_by(area, dept) %>%
  summarize(minimum = min(sale), maximum = max(sale))
```


















